---
title: "Research Question 2"
format: pdf
---

```{r}
library(tidyverse)
library(MASS)
library(caret)
library(psych)
library(ggrepel)
library(MLmetrics)
coffee <- read_csv("../data/coffee_clean.csv", show_col_types = FALSE)
core <- c("aroma", "flavor", "aftertaste", "acidity", "body", "balance", "country_of_origin")
pred_core <- c("aroma", "flavor", "aftertaste", "acidity", "body", "balance")
visualizing <- c("aroma", "flavor", "aftertaste", "acidity", "body", "balance", "country_of_origin", "total_cup_points")

coffee_clean <- coffee %>%
  dplyr::select(all_of(core)) %>% 
  drop_na() %>% 
  mutate(country_of_origin = factor(country_of_origin))

coffee_visualizing <- coffee %>% 
  dplyr::select(all_of(visualizing)) %>% 
  drop_na() %>% 
  mutate(country_of_origin = factor(country_of_origin))
```

\newpage

##

```{r}
summary(coffee_clean)

head(coffee_clean)
```

```{r}
coffee_counts <- coffee_visualizing %>% 
  group_by(country_of_origin) %>% 
  summarise(n = n(), mean_points = mean(total_cup_points, na.rm=TRUE))

ggplot(coffee_counts, aes(x = reorder(country_of_origin, n), y=n, fill=mean_points)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Coffee Cups Tested by Country",
    fill = "Mean Cup Points"
  ) +
  xlab("Number of Cups") +
  ylab("Country of Origin")
```

\newpage

## MANOVA

```{r}
# Fit MANOVA
Y <- as.matrix(coffee_clean[, pred_core])
manova_fit <- manova(Y ~ country_of_origin, data = coffee_clean)

summary(manova_fit, test = "Pillai")

summary(manova_fit, test = "Wilks")

summary(manova_fit, test = "Hotelling-Lawley")

summary(manova_fit, test = "Roy")

summary.aov(manova_fit)
```

\newpage

## LDA

```{r}
# Fit LDA
lda_fit <- lda(country_of_origin ~ ., data = coffee_clean)

lda_fit$svd^2 / sum(lda_fit$svd^2)

lda_pred <- predict(lda_fit)

lda_df <- data.frame(
  LD1 = lda_pred$x[, 1],
  LD2 = lda_pred$x[, 2],
  country = coffee_clean$country_of_origin
)

manual_colors <- c(
  "Brazil"                        = "#1f77b4",
  "Colombia"                      = "#ff7f0e",
  "Costa.Rica"                    = "#2ca02c",
  "Ethiopia"                      = "#d62728",
  "Guatemala"                     = "#9467bd",
  "Honduras"                      = "#8c564b",
  "Mexico"                        = "#e377c2",
  "Taiwan"                        = "#17becf",
  "Tanzania..United.Republic.Of"  = "#bcbd22",
  "Thailand"                      = "#000000",
  "United.States..Hawaii."        = "#ff9896"
)

coffee_lda_plot <- ggplot(lda_df, aes(x = LD1, y = LD2, color = country)) +
  geom_point(alpha = 0.6, size = 2) +
  stat_ellipse(level = 0.68) +
  labs(title = "LDA: Coffee Origins in Discriminant Space",
       x = "LD1", y = "LD2") +
  scale_color_manual(values = manual_colors, name = "Country of Origin") +
  theme_minimal()

ggsave(
  filename = "../figures/07_coffee_lda_plot.png",
  plot = coffee_lda_plot,
  width = 8,
  height = 5,
  dpi = 300)
```

```{r}
# Loadings plot
loadings <- as.data.frame(lda_fit$scaling)
loadings$variable <- rownames(loadings)

lda_loadings <- ggplot(loadings, aes(x = LD1, y = LD2, label = variable)) +
  geom_segment(aes(x = 0, y = 0, xend = LD1, yend = LD2),
               arrow = arrow(length = unit(0.2, "cm"))) +
  geom_text(nudge_y = 0.05) +
  labs(title = "LDA Loadings") +
  theme_minimal()

ggsave(
  filename = "../figures/09_lda_loadings.png",
  plot = lda_loadings,
  width = 8,
  height = 5,
  dpi = 300)
```

```{r}
# Variance explained plot

var_explained <- lda_fit$svd^2 / sum(lda_fit$svd^2)
ve_df <- data.frame(
  LD = paste0("LD", seq_along(var_explained)),
  variance = var_explained
)

ggplot(ve_df, aes(x = LD, y = variance)) +
  geom_col(fill = "#1f77b4") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Variance Explained by Discriminant Axes",
       y = "Proportion of Variance") +
  theme_minimal()
```

```{r}
ve_df$cumulative <- cumsum(var_explained)

lda_variance_explained <- ggplot(ve_df, aes(x = factor(LD, levels = LD))) +
  geom_col(aes(y = variance), fill = "#1f77b4", alpha = 0.7) +
  geom_line(aes(y = cumulative, group = 1), color = "darkred", linewidth = 1) +
  geom_point(aes(y = cumulative), color = "darkred", size = 2) +
  geom_hline(yintercept = 0.90, linetype = "dashed", color = "gray50") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Variance Explained by Discriminant Axes",
       x = NULL, y = "Proportion of Variance") +
  theme_minimal()

ggsave(
  filename = "../figures/08_lda_variance_explained.png",
  plot = lda_variance_explained,
  width = 8,
  height = 5,
  dpi = 300)
```


```{r}
coffee_clean <- coffee_clean %>%
  mutate(country_of_origin = factor(make.names(country_of_origin)))

set.seed(42)
ctrl <- trainControl(method = "cv", number = 10,
                     classProbs = TRUE,
                     summaryFunction = multiClassSummary,
                     savePredictions = TRUE)  # change "final" to TRUE

lda_caret <- train(country_of_origin ~ .,
                   data = coffee_clean,
                   method = "lda",
                   trControl = ctrl,
                   preProcess = c("center", "scale"))

true_levels <- levels(coffee_clean$country_of_origin)
preds <- factor(lda_caret$pred$pred, levels = true_levels)
obs   <- factor(lda_caret$pred$obs,  levels = true_levels)

cm <- confusionMatrix(preds, obs)
cm$byClass[, c("Precision", "Recall", "F1")] %>%
  colMeans(na.rm = TRUE)
```
```{r}
print(lda_caret)
lda_caret$results
```

```{r}
fold_results <- lda_caret$resample

ggplot(fold_results, aes(x = Resample, y = Accuracy)) +
  geom_col(fill = "#1f77b4", alpha = 0.8) +
  geom_hline(yintercept = mean(fold_results$Accuracy), 
             linetype = "dashed", color = "red") +
  labs(title = "Accuracy per CV Fold",
       x = "Fold", y = "Accuracy") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
pr_df <- as.data.frame(cm$byClass[, c("Precision", "Recall", "F1")])
pr_df$country <- gsub("Class: ", "", rownames(pr_df))

pr_long <- tidyr::pivot_longer(pr_df, cols = c(Precision, Recall, F1),
                                names_to = "Metric", values_to = "Value")

ggplot(pr_long, aes(x = reorder(country, Value), y = Value, fill = Metric)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Per-Class Precision, Recall, F1",
       x = "Country", y = "Score") +
  theme_minimal()
```

```{r}
cm_df <- as.data.frame(cm$table)

confusion_matrix <- ggplot(cm_df, aes(x = Prediction, y = Reference, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), size = 2.5) +
  scale_fill_gradient(low = "white", high = "#1f77b4") +
  labs(title = "Confusion Matrix (10-Fold CV Predictions)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


ggsave(
  filename = "../figures/010_confusion_matrix.png",
  plot = confusion_matrix,
  width = 8,
  height = 5,
  dpi = 300)
```


## Presentation

### Question
- Can we predict the country of origin of coffee based on its sensory characteristics?
- Helps us determine if the region causes a difference in the taste of the coffee
- Something useful for coffee brands/distributors when determining where to get their coffee
- Utilize aroma, flavor, aftertaste, acidity, body, balance to determine the country of origin

### MANOVA
- compares the group means, determines if they differ at all
- Null hypothesis: are the group means of all the groups the same?
- Utilize 4 test statistics: Wilks' Lambda, Pillai's Trace, Hotelling-Lawlay, Roy's Largest Root
  - Likely see that Pillai's Trace will be best for the data because the covariance is likely not the same between the groups
  
### MANOVA Results

### LDA
- We see that the groups differ because of the MANOVA results
- LDA helps us illustrate the difference between the groups, and how separable the groups are
- LDA works well because we are utilizing continuous data

### LDA Results

### Cross Validation Performance
- CV is good because it allows us to see how LDA works on unseen data, with 10 fold doing it well for the entire dataset
